---
title: "Standard Statistical Hypothesis Tests in R"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    df_print: paged
runtime: shiny_prerendered
description: >
  Hypothesis tests in R.
---


```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(janitor)
library(DT)
library(knitr)
tutorial_options(exercise.timelimit = 10)

BeanSurvey<-read.csv("bean_survey.csv")[,-1]

BeanSurvey$AGEHH<-c(32,57,20,55,78,42,28,49,41,54,65,63,32,29,43,68,53,39,25,44,23,47,32,26,25,29,44,23,58,69,26,65,75,51,38,24,35,29,37,45,50,23,60,32,70,43,51,28,65,33)


BeanSurvey<-mutate(BeanSurvey,GENDERHH=ifelse(is.na(GENDERHH),"male",as.character(GENDERHH)))

```


## Overview

Doing the standard statistics hypothesis tests in R is pretty easy! In this workbook we are going to look at how you can do some very commonly used tests - t-test and chi-square test, as well as the non-parametric equivalents which you can use in cases when the assumptions are not met. We are going to assume that you have probably come across these methods before, but you also may want to follow some of the linked resources for a refresher.

In general these simple tests can be a useful starting point, but are limited in what they can tell you. So in the next module of the course we will talk more about statistical modeling, which let's you extend upon the principles from these statistical tests and build on them to be applicable in more scenarios and can be applied in much more interesting ways than just these simple tests.

We are going to use the same `BeanSurvey` dataset we looked at earlier in the course to make graphs & data manipulations and we will now extend this into some statistical tests. The data is embedded at the end of this workbook if you want a refresher on what this data contains, and how the variables are named.


There are lots of resources walking through the basic mechanics of the t-test in R. There is a nice guide [here](https://uc-r.github.io/t_test#twosample)

As you will see, the functions themselves are pretty straightforward. And, you may be pleased to hear, there are not very many new functions to learn about in this module, especially compared to the previous modules!

However, it is extremely important to use these functions in the correct context and know what to do when results are unexpected. One of the best things about R is the ability to create a reproducible workflow using markdown files to store commands, and explanations. And in the videos for this module we talk through one such example - trying to emphasise the process of thinking about what you are doing, and what it means, and how it can be improved. 
We will learn more about extending this approach from hypothesis tests into more complete statistical models in the next module of this course.

![](https://youtu.be/LVcmN2OSXY8)

The second part of the video will come later in the workbook!

## t tests

We can use a two sample t-test when we want to compare the means of a numeric variable across two groups, to determine if the difference is statistically significant.

So, if we think back to our `BeanSurvey` dataset - we could compare whether the average amount of land owned by male headed households is different to the average amount of land held by female headed households. We should always start by exploring our data first! 

### Preliminary Analysis

Before any formal hypothesis test, always look at summary statistics and plots showing the relationship you are interested in.

And this is the perfect chance to apply and recap all your newly acquired ggplot2 and dplyr skills! 

*In the code box below write the code to match the embedded output, to show the means and standard deviations of `LANDAREA` by `GENDERHH`, and the number of observations in each group*


```{r,echo=FALSE,message=FALSE}
BeanSurvey %>% 
   group_by(GENDERHH) %>%
    summarise(n=n(),mean=mean(LANDAREA,na.rm=T),sd=sd(LANDAREA,na.rm=T))
```


**Summary Statistics of Land Area by Gender of Household head**
```{r summarystats, exercise=TRUE,error=TRUE}

```


```{r summarystats-solution}
BeanSurvey %>% 
   group_by(GENDERHH) %>%
    summarise(n=n(),mean=mean(LANDAREA,na.rm=T),sd=sd(LANDAREA,na.rm=T))
```

*Now match embedded output, this time to show histograms and boxplots of `LANDAREA` split by `GENDERHH`*
**Histograms of Long Rains Yield by VILLAGE**
```{r,echo=FALSE,message=FALSE}
BeanSurvey %>% 
  ggplot(aes(x=LANDAREA))+
    geom_histogram()+
      facet_wrap(~GENDERHH)

BeanSurvey %>% 
  ggplot(aes(y=LANDAREA,x=GENDERHH))+
    geom_boxplot()
```

```{r summaryplot, exercise=TRUE}

```

```{r summaryplot-solution}
BeanSurvey %>% 
  ggplot(aes(x=LANDAREA))+
    geom_histogram()+
      facet_wrap(~GENDERHH)

BeanSurvey %>% 
  ggplot(aes(y=LANDAREA,x=GENDERHH))+
    geom_boxplot()
```

Looking at the summary stats we can see that the average values are a little higher for male headed households on average as compared to female headed households.

We can see from the boxplots and histograms that generally the distributions of landarea are shifted a little higher for the male headed households. The medians and quartiles making up the boxes are higher for male headed households. However, we also see that the distribution is skewed by some large values - most farmers have under 3 acres of land, but there are a small number with much larger yields. 


So from the plots and summary statistics it looks like male headed households have, on average, more land than female headed households. We can conduct a formal statistical significance test to see if there is sufficient evidence to confirm that hypothesis, or whether the observed differences in the data are no larger than those we would expect to see by random chance.

### Running the t-test

We can use a t-test when we have a continuous numeric variable and we either want to compare whether the average value is significantly different from a fixed reference value (a one sample t-test) or whether the average value is significantly different between two groups (a two sample t-test). In this case we have appropriate data to use - a numeric variable `LANDAREA` and a two-level categorical variable `GENDERHH`.

The structure of the `t.test` function is that we provide the name of our numeric variable (`LANDAREA`), followed by a tilde (`~`), followed by the name of our 2 level grouping variable (`GENDERHH`). Then we need the name of our data. 

```{r ttest1, exercise=TRUE}
t.test(LANDAREA~GENDERHH,data=BeanSurvey)
```

In this case we would conclude, from the p-value of 0.2496 that there is not enough evidence to say that there is a significant difference in the mean values of land area based on gender of head of household.

The output also shows us the t-statistic, degrees of freedom and a 95% confidence interval around the difference in the means.

It's worth reminding yourself of how p-values work - they are very commonly misunderstood and therefore misinterpreted. There is a great video from Dr Nic about p-values [here](https://www.youtube.com/watch?v=eyknGvncKLw)


With the `t.test` function it is important to note that, unlike most of the functions we have used so far in this course the name of the data comes as the second argument *after* the formula. 

We can use a pipe (`%>%`) to get from the data to the t-test, but we need to use it in a slightly different way. The pipe assumes the first argument of a function is the data, but we can override this. Within `t.test`, or any function where data is not the first argument, we need to include an argument `data= .` in a line of code following a pipe `%>%`. The `.` denotes to inherit the data coming in from the previous line.


```{r pipein, exercise=TRUE}
BeanSurvey %>% 
     t.test(LANDAREA~VILLAGE,data=.)

```
And that's it! We shown you how to do a t-test in R...

... but! There are also other more practical things to consider when running a t-test; whether the hypothesis and the assumptions underlying it are sensible. And we should know what to do if we do encounter anything that makes us question the validity of the methods.

There is actually one assumption you were probably taught about when learning the t test that you do not need to worry about at all when using the `t.test` function in R.

### Assumption 0: Equal Variance 

The result you obtain by default from the `t.test` function is not exactly the same as you would see in other software packages, because R has slightly different default settings.

You probably learnt that you need to have approximately "equal variance" within your two groups for a t-test to be valid. However, by default, R uses a "Welch" t-test. This is a modification to the classic t-test which does not require equal variance. 

There is no real reason not to use this modification. If the variances are exactly equal then both the classic and Welch t-tests will provide identical results; and when the variances are very similar the differences between the results are trivial.

But if the variances are not similar the classic t-test results would not be valid where the Welch t-test results would be. Looking at the summary statistics and plots we produced earlier it does appear that the variances may not be equal - the standard deviation is quite a bit larger for male headed households compared to female headed households. So in this case we would almost certainly want to keep with the default from R and use the Welch version of the t-test.

But, if you really love the classic t-test, or just want to confident that the results match those you might have seen running the same code from other software  then you can add the argument `var.equal = TRUE` into the code

```{r ttest, exercise=TRUE}
t.test(LANDAREA~GENDERHH,data=BeanSurvey,var.equal=TRUE)

```

As you can see, if you go back and compare the results to the previous output, the value of the t-statistic, and degrees of freedom change very slightly, and so does the p-value. In this case though, it would have made no substantial difference to our conclusions, despite the fairly large change in the p-value. In other cases it might be the difference between a significant and a non-significant result, so we should think carefully before assuming that the equal variance assumption is valid.

### Assumption 1: Normality

We also have another assumption that we have approximate normality of the response with each group. The key word here though is 'approximate';  a lot of people can get very hung up on whether they have a perfectly normal distributions. In fact, as we increase the size of our data, the t-test will still be valid with increasingly non-normal looking distributions. This is known as the 'central limit theorem'.

Again, Dr Nic, is on call with a video about this [here](https://www.youtube.com/watch?v=_YOr_yYPytM&ab_channel=DrNic%27sMathsandStats)

It is usually worth checking the histogram to check that we have a roughly symmetrical distribution which is unimodal. 

There are formal hypothesis tests we could use to 'test' normality, that some people might teach you about. But these are completely useless within this context.

There is no assumption requiring 'exact' normality. And using the tests of normality, the more data that we have the more likely we are able to identify that something is not 'exactly' normal. This is the exact opposite of what we need to satisfy this assumption of the t-test, because of the central limit theorem!

So simply visually inspecting the distributions from a histogram, is usually sufficient for us when we consider if our data can meet this assumption.

In this particular case we probably do not have enough data to be able to invoke the central limit theorem. And the histograms we produced earlier suggested a very substantial skew in the data. So we should consider possible alternatives. 

### Non-Parametric Alternative: Wilcoxon Test

A common solution to this problem would be to use a non-parametric test, like the Wilcoxon rank sum test, also known as the Mann-Whitney U test.

Instead of testing whether the mean values of land area differ between the groups, this considers whether the distribution of land area is shifted for male heads as compared to females, rather than whether the mean value is different. 

This works by ranking the values in each group and comparing the average ranks in the two groups. This makes it robust to all sorts of strange looking distributions, because the maximum value can never be more than one unit higher than the next highest value. 

However, it does have less statistical power to detect differences between groups than a t-test, which is why the t-test is often preferred.

Luckily, within R, the code for `wilcox.test` function is almost identical to what we have seen already with `t.test`. So switching to a non-parametric test is simple! We just change `t.` to `wilcox.`.

```{r wilcox1,exercise=TRUE}
  wilcox.test(LANDAREA~GENDERHH,data=BeanSurvey)
```

And here we obtain a similar result with a p-value of p=0.3469 suggesting we have no evidence against the null hypothesis that distribution of land area for male heads is the same as for female heads. So this provides further reinforcement of our initial conclusion from the t-test.  

The hypothesis is described in the output as "location shift" - this is another way of considering the 'mean of the ranks'. 

Because of the existence of ties, the Wilcoxon test does not directly compare the median values, which is a common misconception. In the video, you will have seen an example where the medians of the two groups are the same but the Wilcoxon test provides a significant result.

You can see a better idea of what the Wilcoxon test is actually doing by exploring a little using the `rank()` function to rank the values and then calculate the mean of the ranks and then pipe into a plot or summary statistics.

The `rank()` function is quite simple, if we provide a vector of numbers it will give rank 1 to the lowest, 2 to the next lowest and so on.

```{r ranksimple,exercise=TRUE}

some_numbers<-c(45,1,100,8)

some_numbers
rank(some_numbers)

```

So we can add the ranks of `LANDAREA` into a new column of our data using `mutate` and then investigate the summary statistics on this new rank variable to get a better understanding of the comparison by made in the Wilcoxon test.

```{r rank2,exercise=TRUE}
BeanSurvey %>%
  select(LANDAREA,GENDERHH) %>%
  mutate(land_rank=rank(LANDAREA))
```

```{r rank5,exercise=TRUE}
BeanSurvey %>%
  select(LANDAREA,GENDERHH) %>%
  mutate(land_rank=rank(LANDAREA)) %>%
    ggplot(aes(y=land_rank,x=GENDERHH))+
      geom_boxplot()
```


```{r rank3,exercise=TRUE}
BeanSurvey %>%
  select(LANDAREA,GENDERHH) %>%
  mutate(land_rank=rank(LANDAREA)) %>%
    group_by(GENDERHH) %>%
      summarise(mean(land_rank))
```

The smallest value is ranked as 1, the second smallest value ranked as 2 and so on. When we compare the distribution of ranks we can see the ranks are a little higher (i.e. larger area) for the male headed households, but again the difference is not large enough to conclude that this difference is larger than what we would see by chance.

If you haven't come across it before, you can read more about the Wilcoxon rank sum test [here](https://www.stat.auckland.ac.nz/~wild/ChanceEnc/Ch10.wilcoxon.pdf). 

### Assumption 2: Independence

Probably the most important assumption to consider is whether or not the observations are independent of each other in the two groups. The simplest example where this is not the case when we have paired data. In that scenario we would use a paired t-test, and this can also be conducted using `t.test()`. However we apply this function in quite a different way when we carry out the paired t-test, compared to the previous example we saw, when we applied an independent two-sample t-test.


A good example of this would be if we were to compare beans planted in the short rains against beans planted in the long rains. We can produce separate histograms or boxplots for the beans being planted in long rains, and beans being planted in short rains, using seperate calls to `ggplot()`. Because they are in different variables we cannot directly map them onto the same plot like we have seen before - we would need to do some additional data manipulation first. We will learn more about that in the final module of the course, which will make it a bit tidier, but producing two separate plots is still an effective way of exploring the data here.  

```{r pair1,exercise=TRUE ,echo=FALSE}
ggplot(data=BeanSurvey,x=BEANSPLANTED_LR)+
  geom_histogram()

ggplot(data=BeanSurvey,x=BEANSPLANTED_SR)+
  geom_histogram()
```

And again, when we calculate mean and standard deviation values, we do not need to use group_by to see the differences between beans planted in long rain season and beans planted in short rain season. Since all the households have values for both variables, it’s not the households that define the groups, but the choice of variable. 

```{r summ_groups, exercise=TRUE}
BeanSurvey %>%
  summarise(LR_mean=mean(BEANSPLANTED_LR,na.rm=T),LR_sd=sd(BEANSPLANTED_LR,na.rm=T),
            SR_mean=mean(BEANSPLANTED_SR,na.rm=T),SR_sd=sd(BEANSPLANTED_SR,na.rm=T))
```

When it come to now carrying out the hypothesis test through a paired t-test, we using the same `t.test` function, but with a slightly different way of specifying the inputs. 

```{r paired, exercise=TRUE}
t.test(BeanSurvey$BEANSPLANTED_LR,BeanSurvey$BEANSPLANTED_SR,paired=TRUE)
```
In this case although we can see that on average 1.3kg more beans were planted in the long rains, we do not have enough evidence at the 5% significance level to conclude that this difference is statistically significant (p=0.0978).

Note that this time, it is not possible to pipe into the paired t-test since none of the arguments are `data`, instead we need two supply two individual variables names. Lots of functions work with pipes, but not all!

Again, we may be concerned about the normality assumption. With the paired t-test what we care about is not the distribution of the individual variables but the distribution of the differences within each household So we can use `mutate` to calculate a column of differences and then plot a histogram of the differences.

```{r pairhist, exercise=TRUE}
BeanSurvey %>%
  mutate(PLANTED_DIFF=BEANSPLANTED_LR-BEANSPLANTED_SR) %>%
    ggplot(aes(x=PLANTED_DIFF))+
      geom_histogram()
```

We may be concerned that results are being skewed by the two very high values, and again we don't have a very large data set so we cannot quite rely on the central limit theorem. So perhaps we could also try the Wilcoxon test again. And, luckily this works exactly like before following a very similar pattern. The `paired` code for `wilcox.test` is the same as the `paired` code for `t.test`, all we need to do is change the function.


```{r pairwilcox, exercise=TRUE}
wilcox.test(BeanSurvey$BEANSPLANTED_LR,BeanSurvey$BEANSPLANTED_SR,paired=TRUE)
```
Again, we obtain a similar conclusion, we do not have enough evidence to be confident that the difference between the seasons in the data is due to more than just chance.

The paired t-test works in this specific scenario, but there are more complicated violations of independence, which can't be so easily fixed. 

Particularly when we consider having more than two matched observations (e.g. data taken over multiple time points); experimental designs with blocking; or clustered sampling in surveys. 

In these cases the t-test is not at all a sufficient method for analysis! So, you will have to start learning more about statistical modeling to do a good job of analysing your data - and matching the appropriate model structure to your data structure.



## Chi-square test

![](https://youtu.be/a-hVoVnkWSg)

Another common, simple statistical test is the chi-square test. We use this when we want to compare whether or not there is a relationship between two categorical variables.

In this workbook I am going to demonstrate using the `janitor` library for this sort of analysis. This allows you to produce tables, percentages, and chi-square tests in the same tidyverse style as `dplyr`.

As this is quite a new package, when you look online there are not many resources which take this particular approach, and you will instead see other functions, like `table` or `prop.table` being used. There is always more than one way to do something in R. For me the `janitor` approach is actually much more straightforward to follow than the traditional approach.

So, as an example, we may be interested in the relationship between whether a household sells beans, `SELLBEANS`, and the village `VILLAGE`.

### Preliminary analysis

The preliminary analysis that we would first look at here would be a cross-tab comparing these variables.  We can pipe `%>%` directly from the data into the `tabyl` function, and then specify the variables to include in our cross-tab.

```{r tabyl1,exercise=TRUE}

BeanSurvey %>%
  tabyl(VILLAGE,SELLBEANS) 

```

The first variable will always be placed in the rows of the table, and the second variable will be placed in the columns.

Try to interpret those numbers from the frequency table output, and see if you think there is a pattern. It can sometimes be hard, especially if you have very uneven group sizes.

So we often will also look at marginal percentages. That: is for each level of one variable what percentage of respondents fall in the levels of the second variable. 

We refer to these as "row percentages" or "column percentages", depending on whether we are looking at the percentages within each row or percentages within each column. We can pipe the frequency table into the function `adorn_percentages` to investigate this.

I.e. We have two possible questions:  "What percentage of people in Kimbugu sell beans?" This is the 'row percentage' because we are looking within each village levels, and `VILLAGE` is the row factor. The row percentage is the default option, so we don't need to add anything else to this function.

```{r tabyl2,exercise=TRUE}

BeanSurvey %>%
  tabyl(VILLAGE,SELLBEANS) %>%
    adorn_percentages()
```
By default this shows us proportions rather than nicely formatted percentages. But we can pipe into an extra function, `adorn_pct_formatting` to apply the percentage formatting.

```{r tabyl3,exercise=TRUE}
BeanSurvey %>%
  tabyl(VILLAGE,SELLBEANS) %>%
    adorn_percentages() %>%
      adorn_pct_formatting()
```

So we can see that, 54% of people in Kimbugu sell beans, and only 15% of people in Lwala sell beans.

Or we could ask for "What percentage of people who sell beans live in Kimbugu". We need to add the argument "col" into the `adorn_percentages` function to obtain this, as it is called the "column" percentage. We are looking within the column factor, `SELLBEANS`

```{r tabyl4,exercise=TRUE}
BeanSurvey %>%
  tabyl(VILLAGE,SELLBEANS) %>%
    adorn_percentages("col") %>%
      adorn_pct_formatting()
```

So here we can see that 77% of those who sell beans live in Kimbugu.

This is not a contradiction! 

We always have to think carefully about which percentages are being calculating, and thinking about which would make the most sense to interpret. 

In this case, the row percentages would be likely much more helpful to us than the column percentages. The more interesting question, and probably easier to interpret, would be to compare relative levels of bean selling across the different villages

To make a plot of these results, we would probably want to produce some bar charts. There are a few different ways of making bar charts we learnt about earlier in the course - either stacking or using facets. 

*See if you can replicate the plot below, or choose to produce a different graph which shows the relationship between `VILLAGE` and `SELLBEANS`. Make sure to modify the scale on the x axis as well to make it more readable!*
```{r ggbar0,echo=FALSE}
ggplot(data=BeanSurvey,aes(group=SELLBEANS,fill=SELLBEANS,y=VILLAGE))+
  geom_bar(position="fill")+
   scale_x_continuous(breaks=c(0,0.2,0.4,0.6,0.8,1),labels=c("0%","20%","40%","60%","80%","100%"))
```

```{r ggbar1,exercise=TRUE}

```


```{r ggbar1-solution}
ggplot(data=BeanSurvey,aes(group=SELLBEANS,fill=SELLBEANS,y=VILLAGE))+
  geom_bar(position="fill")+
   scale_x_continuous(breaks=c(0,0.2,0.4,0.6,0.8,1),labels=c("0%","20%","40%","60%","80%","100%"))
```

### Running the test

To run the chi square test, we need to have the two-way frequency table exactly as it comes from running the `tabyl()` line of code. This then pipes directly into the `chisq.test` function.

```{r chi1,exercise=TRUE}
BeanSurvey %>%
  tabyl(VILLAGE,SELLBEANS) %>%
    chisq.test()
```

And we obtain a p-value of 0.0095, saying that here we have enough evidence to conclude that farmres in Kimbugu are more likely to sell their beans compared to farmers in Lwala.


Make sure you pipe from the frequency table, created by `tabyl`, into `chisq.test` and not from the table of proportions or formatted percentages, created by either `adorn_percentages` or `adorn_pct_formatting`. 

Piping from the table of proportions *does not give an error* but the results it returns *are nonsense*.

```{r chi1_fail,exercise=TRUE}
BeanSurvey %>%
  tabyl(VILLAGE,SELLBEANS) %>%
     adorn_percentages() %>%
        chisq.test()
```
Do not do this!


### Assumptions

There are two key assumptions for a chi-square test to be valid.

The first is similar to the t-test, and the Wilcox test. We need independent observations. If we don't meet this assumption, we will have to learn more about statistical modelling, rather than use a chi-square test.


There is another assumption we have to meet though related to the sample size within each category.
Let's try a different question - whether `HHTYPE` is related to the selling of beans. See if you can match the output below:


```{r chi_q1_show,echo=FALSE}
BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) 

BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) %>%
     adorn_percentages()

BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) %>%
    chisq.test()
```

```{r chi_q1,exercise=TRUE}

```

```{r chi_q1-solution}
BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) 

BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) %>%
     adorn_percentages()

BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) %>%
    chisq.test()
```
We need to have a sufficient number of observations in all possible combinations of the categories of our two variables - in this example we see a warning message that the "chi-square approximation may be incorrect". The general rule of thumb is to have expected frequencies of at least five observations in all combinations - we can clearly see many of the household types have fewer observations than that overall, before we even consider whether they sold beans or not. 

In this case, we have three options for how to deal with it:  
1. Use a different variable - like gender of household head. In this case we have it already in our data, but if we did not we could create this using `mutate` and then the `recode` function which is very useful to know about.

In the recode function we first provide the variable we are recoding, then we specify how we want to recode it following the pattern `"existing value" = "new value"`. So any of the female headed categories I recode into the same new category. Looking into the data I discovered that the "other" household was male headed. But if I was recoding the data without knowing this, it is probably a more sensible move to choose to remove this value using a `filter`.

```{r chi_q2,exercise=TRUE}
BeanSurvey %>%
  mutate(HHTYPE2 = recode(HHTYPE,
                          "Female headed absentee husband"="Female headed",
                          "Female headed, no husband"="Female headed",
                          "Male headed more than one wife"="Male headed",
                          "Male headed one wife"="Male headed",
                          "Single man"="Male headed")) %>%
  filter(HHTYPE2!="Other") %>%
  tabyl(HHTYPE2,SELLBEANS)


```

When using `janitor` and the `tabyl` functions it may often make sense to save the initial `tabyl` as an object. This way we can then move to the subsequent steps, of looking at percentages and conducting hypothesis tests, without replicating the data manipulation steps every time

```{r chi_q99,exercise=TRUE}
tabyl1<-BeanSurvey %>%
  mutate(HHTYPE2 = recode(HHTYPE,
                          "Female headed absentee husband"="Female headed",
                          "Female headed, no husband"="Female headed",
                          "Male headed more than one wife"="Male headed",
                          "Male headed one wife"="Male headed",
                          "Single man"="Male headed")) %>%
  filter(HHTYPE2!="Other") %>%
  tabyl(HHTYPE2,SELLBEANS)


tabyl1

tabyl1 %>%
  adorn_percentages()

tabyl1 %>%
    chisq.test()
```



2. Only include the two common `HHTYPES` which are "Female headed, no husband", and "Male headed one wife". We could restrict this using `filter`

```{r chi_q3,exercise=TRUE}
tabyl2 <- BeanSurvey %>%
  filter(HHTYPE%in%c("Female headed, no husband","Male headed one wife")) %>%
  tabyl(HHTYPE,SELLBEANS)

tabyl2

tabyl2 %>%
  adorn_percentages()

tabyl2 %>%
    chisq.test()
```
In this case we still do get the same warning - because the frequencies in the "female headed, no husband" category are still a little small.


3. Use the non-parametric alternative, the Fisher Exact Test

Thankfully, much like with the `wilcox.test()` function, we can use a non-parametric test when dealing with two categorical variables by just changing the function name from `chisq.test()` to `fisher.test()`.

```{r chi_q4,exercise=TRUE}
BeanSurvey %>%
  tabyl(HHTYPE,SELLBEANS) %>%
    fisher.test()
```
Whichever method we used here we reached the same conclusion - no evidence of relationship between household type and selling beans. But notice that the hypotheses for options 1 and 2 are different. In option one we are comparing differences only in the gender of the head, not consdiering any of the other household composition that exists in the `HHTYPE` variable. And in option 2 we are only comparing the two specific household types.

Unlike with the `t-test()` we do not change the hypothesis we are testing by using the Fisher test instead. However this test is a lot more computationally intensive. With even a reasonably small dataset (200 or so observations) you may find that it takes some time for the computer to return the result. This is why the chi-square test is often preferred, since it is always very quick for the computer to calculate!



## More tests?

These are the only hypothesis tests we will talk about explicitly in this course, but hopefully you will see that the syntax for producing these tests is not especially complicated. This is true for all commonly used statistical hypothesis tests. 

If you know the simple statistical test you would like to conduct, and your data is in the appropriate format for that test, and the assumptions for that test can be met, it is usually pretty simple to find and then write the R code to carry out the test. There are a lot of statistical tests within base-R!

```{r,echo=FALSE}
apropos("\\.test")
```

And those are just the tests which are in base-R - there will be many more in various additional libraries!

But 'getting data in the appropriate format', 'checking the assumptions for the test' and 'knowing whether the hypothesis test makes sense for your question' are not so straightforward! It is incredibly easy to do terrible data analysis in R, by skipping straight to the end p-value and churn out meaningless p-values. It is a bit like driving a car - anyone can push the accelerator and make it go forward; but if you don't know how to change the gears or you completely ignore all of the road signs then you are probably going to crash.

This is, of course, not just true of R but of any analysis software. However with R there is even less of an excuse for doing this since all of the tools needed to prevent it are provided and freely available. This is why we have placed so much emphasis in this course on data manipulation (`dplyr`), exploratory graphics (`ggplot2`) and using a coherent and reproducible workflow (RStudio). 

When doing research in the real world, it is also extremely unlikely that a simple hypothesis test will be sufficient for your analysis. But these are often what you are taught in your statistic modules, because they are the first step into more widely applicable statistical models. It is likely you will need to be doing some statistical modelling, to better explain the patterns and trends in your data. We will learn about this in the next module!

## Exercises

In this module the exercises are a little different. It is now up to you to use RStudio to complete the exercises on your own computer.

You can download the files you need [by clicking on this link here](https://github.com/stats4sd/R4CCRP_06_StatsTest/blob/main/Module%206%20Exercises.zip?raw=true)


Make sure you unzip these into a new folder, and then start a new project file based on that folder, like we learnt in the previous module.

You may need to install any of the packages used before proceeding - make sure you have these installed and loaded, by running the first code chunk, before you try to start solving any of the questions!

## References  

Video on interpreting p-values (Dr Nic's Maths and Stats):  
https://www.youtube.com/watch?v=eyknGvncKLw  

Video on choosing an appropriate test (Dr Nic)
https://www.youtube.com/watch?v=rulIUAN0U3w

Tutorial on t-tests and non-parametric equivalents (UC Business Analytics R Programming Guide):  
https://uc-r.github.io/t_test  

Tutorial covering all the tests covered in this session in more detail. (Note - Part I (Chi square tests) is much more complicated in this tutorial than the method we have shown! Suggest only working through Parts II and III)
https://sbc.shef.ac.uk/workshops/2019-09-18-stats-r/practical.nb.html


## Appendix: 'BeanSurvey' dataset 

The data we are using in this session is an extract of a survey conducted in Uganda from farmers identified as growing beans.

The dataset contains an extract of 50 responses to 23 of the survey questions, and has been imported to R as a data frame called `BeanSurvey`.

A summary of the columns in the dataset is below.

```{r, echo=FALSE,message=FALSE,warning=FALSE}


data.frame(Column=colnames(BeanSurvey),
           Description=c("Farmer ID","Village name","Household composition","Gender of Household Head","Age of Household Head",
                         "Occupation of Household Head","Number of Adults within the household","Number of Children (<18) within the household","Do they grow matoke?",
                         "Do they grow maize?","Do they grow beans?","Do they grow banana?","Do they grow cassava?",
                         "Do they grow coffee?","Land area of farm (acres)","Labor usage","Intercrops with beans",
                         "Household decision responsibility","Do they grow beans for sale?","Quantity of beans planted in long rain season",
                         "Quantity of beans planted in short rain season","Quantity of beans harvested in long rain season"
                         ,"Quantity of beans harvested in short rain season")) %>% kable()

```


Spend some time looking through the exploring the full dataset embedded below, to familiarise yourself with the columns and the type of data stored within each column. You may need to refer back to this data at times during this tutorial. Remember that R is case sensitive, so you will always have to refer to the variables in this dataset exactly as they are written in the data. There is a column in this data called "GENDERHH" but there is no column in this data called "GenderHH".

```{r,echo=FALSE}
DT::datatable(BeanSurvey)
```

(You can use the arrow keys on your keyboard to scroll right in case the data table does not fit entirely on your screen)
